A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-10 paper.tex
Pass at intro

\textbf{Keywords:} Ancestral Recombination Graphs

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of a
phylogenetic tree is a necessary prerequisite for many analyses of
species~\citep{rannala2003genetics}, non-recombining sections of
DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
and viruses~\citep{grenfell2004science}.
A phylogenetic tree captures information about the genealogy
of the sample in an elegant and concise
way, by postulating a set of ancestors common to the samples
together with paths of inheritance between them. A rich literature exists
on analysing the mathematical properties of these
trees~\citep{steel2016phylogeny}, and numerous
methods exist to infer them~\citep{felsenstein2004inferring}.
However, the genetic genealogy of a recombining organism
cannot be captured by a single tree: instead,
the inheritance of each section of the genome may be described
by a (slightly) different local tree~\citep{hudson1983properties}.
Inferring complex
genealogies of this nature
presents profound technical difficulties, meaning genealogy-based approaches have
not been widely used in population and statistical genetics.
Until recently, analysis of
population genetic data has therefore primarily focussed on summaries derived
from the local trees~\citep{tajima1983evolutionary,tavare1984line}
rather than using the full underlying genealogy itself. Such
derived measures include allele frequencies and site frequency
spectra~\citep{achaz2009frequency,ralph2020efficiently},
patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
principal components~\citep{mcvean2009genealogical}.

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogy-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as inputs have begun to
appear~\citep{osmond2021estimating,zhang2023biobank,fan2022genealogical,hejase2022deep}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which risks hampering
progress: we currently lack a well-defined data-model and shared terminology
to discuss genealogies in the presence of recombination. This leads to basic errors in statements
about what different inference methods produce as output, as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\item Despite its status as ``the ideal data structure for
population genomic analysis''~\citep{rasmussen2014genome}, there has been
little discussion of its computational properties. We discuss the
relationship between the classical concept of an ARG and the
recently introduced succinct tree sequence data structure.

\item This, coupled with
    the explicit association of genome coordinates with edges, allows
    us to represent even the bewilderingly complex patterns of relatedness
    present in modern datasets.

\end{enumerate}

>>>>

\textbf{Keywords:} Ancestral recombination graphs

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Inferring the detailed genetic genealogy of a set of sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as inputs have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}. [THIS IS BAD, WHY?]
We also lack a shared format for interchanging ARG data [which is bad
for users, progress]. [Because we don't really agree on what an ARG is,
we don't really know what properties it should have, when we infer it].

% What do we do?
Here we precisely define an encoding for ARGs based on specific genomes
and their intervals of genetic inheritance. We contrast this ``gARG''
encoding with the classical approach derived from the coalescent
with recombination stochastic process of Griffiths
and colleagues~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}. As this is based
on recording evolutionary events, we refer to it as the event ARG, or
``eARG'' encoding.
We show that the gARG encoding is both flexible and efficient,
and can represent varying levels of precision about historical
events.
[MORE]

% and because it is focused on encoding the \emph{outcome}
% of evolutionary events rather than the details of the events themselves,

% In this paper we [turn into proper round up paragraph later]
% \begin{enumerate}
% \item Discuss the classical Ancestral Recombination Graph formulation,
% and show how it derives from (and is limited by) the coalescent with
% recombination.
% \item In the following sections we
% show that, although the ARG-as-a-data-structure perspective is
% now the most common interpretation, the \emph{form} of that
% data structure is still closely linked to the original
% stochastic process, and that this form is inflexible
% in the types of genetic inheritance that can be described,
% computationally inefficient, and
% forces complete (and potentially spurious) precision about
% the location and timing
% of recombination events. % (and coalescent events?)
% \item Suggest a definition of ARGs that is free of these limitations,
% where the formulation is derived from a (normally diploid) pedigree rather than from
% coalescent approximations, and where the genetic
% material which is transmitted is described by annotations attached to
% edges rather than nodes.
% \item Suggest a classification of ARG nodes that helps us to understand
% the properties of the structures inferred by different methods, and illustrate
% with some examples.

% \item This, coupled with
%     the explicit association of genome coordinates with edges, allows
%     us to represent even the bewilderingly complex patterns of relatedness
%     present in modern datasets.

% \end{enumerate}
